---
title: Small Telescopes for Higher-Power Replications
author: Lukas wallrich
date: '2025-08-09'
slug: small-telescopes-for-higher-power-replications
categories: []
tags: [Metascience]
subtitle: ''
summary: ''
authors: []
lastmod: '2025-08-09T15:04:49+01:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
projects: []
---

> **TL;DR.** In Simonsohn’s Small‑Telescopes framework you test whether the true effect is **smaller than** the effect that would have given the **original** study 33% power (call this \(d_{33\%}\)). Simonsohn (2015) shows that this requires 2.5x the original sample size if you want 80% power to show that the true effect is smaller than \(d_{33\%}\) *if the true effect is 0*. In this post, I provide alternative multiples for 90% and 95% power (3.5x and 4.5x respectively).


## Small-Telescopes in Short: What is being tested (and why “one‑sided”)?

Let \(d_{33\%}\) be the effect size that would have given the **original** two‑sample *t* test **33% power** at two‑sided \(\alpha_0=.05\). The replication runs a **one‑sided** test

\[
H_0: d = d_{33\%}\quad \text{vs}\quad H_1: d< d_{33\%}
\]

at \(\alpha=.05\), and we plan the replication to have (e.g.) **80% power** to reject \(H_0\) **if the true effect is 0**. *Intuition:* Rather than asking whether the effect is exactly zero, the small telescopes test asks whether it is smaller than the minimal effect the original study had a reasonable chance to detect. If the replication can rule out effects of that size, the onus shifts — the original proponents are suggested to now bear the "burden of proof" again, because even the smallest effect they could plausibly have detected is inconsistent with the replication data.

## When might higher power be worth it?

The original small telescopes rule-of-thumb — about 2.5× the original sample size for 80% power — works well for many replications. But there are situations where aiming higher is justified. For example:

- **Policy- or practice-relevant findings** where acting on a false positive would have high costs.
- **Expensive or rare data** where you won’t get a second chance to replicate, so making the most of the opportunity matters.
- **Adversarial collaborations** where proponents and critics may want to aim for a definitive answer.

In these cases, the (substantial) cost of increasing power from 80% to 90% or 95% may be outweighed by the clarity it brings. A higher-powered replication yields in a more precise effect size estimate, and is more likely to confirm whether the effect is significantly smaller than the small telescope threshold.


## Analytic multipliers

The multipliers for the original per‑cell \(n\) to achieve higher power in the small telescopes test can be derived from the normal approximation of the two‑sample *t* test. Under a normal approximation, the replication‑to‑original per‑cell‑\(n\) ratio is

\[
m=\left(\frac{z_{1-\alpha}+z_{\text{power}}}{z_{1-\alpha_0/2}+z_{p_0}}\right)^2,
\]

with \(p_0=1/3\) (see SM in Simonsohn, 2015).[^1] This gives: **80% → 2.64×**, **90% → 3.66×**, **95% → 4.63×**. With finite‑sample *t* tests, the exact multipliers are **slightly lower** for small samples and approach these limits as \(n_0\) grows. In line with the original paper, which proposed 2.5x for 80% power, I propose using **3.5x** and **4.5x** for 90% and 95% power respectively.[^2]


## Exact calculation and validation of the heuristic multipliers

Using the *t*-distribution for specific sample sizes, we can calculate more specific multipliers for the small telescopes test. The code below computes these "exact" multiplier for a given original per‑cell \(n\) and target power, and compares it to the normal approximation. The results show that the multiples lead to a power within 2% of the target power for \(n_0\) up to 1000 participants per cell.

<details>
  <summary>Show code to calculate exact multiplier</summary>

```{r, message=FALSE, warning=FALSE}

library(dplyr)
library(purrr)
library(pwr)

# d_33% for the original two-sample t (per-cell n), p0 = 1/3 per supplement
d33_two_sample_t <- function(n_per_cell, p0 = 1/3, alpha0 = 0.05) {
  pwr.t.test(n = n_per_cell,
             power = p0,
             sig.level = alpha0,
             type = "two.sample",
             alternative = "two.sided")$d
}

# Power to reject H0: d = d33 with a one-sided (less) t-test at alpha, when true d = 0
# Test is one-sided because the replication has a directional hypothesis
power_reject_d33_under_null0 <- function(n_per_cell_rep, d33, alpha = 0.05) {
  df    <- 2 * n_per_cell_rep - 2
  tcrit <- qt(alpha, df = df, lower.tail = TRUE)
  ncp   <- -d33 * sqrt(n_per_cell_rep / 2)
  pt(tcrit, df = df, ncp = ncp, lower.tail = TRUE)
}

# Exact multiplier via t; returns integer per-cell n for replication and achieved power
small_telescopes_multiplier_exact <- function(n_original_per_cell,
                                              target_power = 0.80,
                                              alpha = 0.05,
                                              p0 = 1/3,
                                              alpha0 = 0.05,
                                              max_mult = 100,
                                              tol = 1e-7) {
  stopifnot(n_original_per_cell >= 3)
  d33 <- d33_two_sample_t(n_original_per_cell, p0 = p0, alpha0 = alpha0)
  f <- function(n_rep) power_reject_d33_under_null0(n_rep, d33, alpha) - target_power

  lower <- max(3, n_original_per_cell) * 1.0
  upper <- min(max_mult * n_original_per_cell, 1e7)
  f_lower <- f(lower); f_upper <- f(upper)
  while (f_lower * f_upper > 0 && upper < 1e7) {
    upper <- min(upper * 2, 1e7); f_upper <- f(upper)
  }

  n_rep_cont <- if (f_lower >= 0) lower else if (f_upper <= 0) upper else
    uniroot(f, interval = c(lower, upper), tol = tol)$root

  n_rep_int <- ceiling(n_rep_cont)
  tibble(
    n_original_per_cell = n_original_per_cell,
    d33 = d33,
    n_replication_per_cell_exact = n_rep_int,
    exact_multiplier = n_rep_int / n_original_per_cell,
    achieved_power_exact = power_reject_d33_under_null0(n_rep_int, d33, alpha)
  )
}

# Normal-approx multiplier (independent of n0) for p0 = 1/3
small_telescopes_multiplier_normal <- function(target_power = 0.80,
                                               alpha = 0.05,
                                               p0 = 1/3,
                                               alpha0 = 0.05) {
  z_alpha0_2      <- qnorm(1 - alpha0/2)
  z_p0            <- qnorm(p0)
  z_1_minus_alpha <- qnorm(1 - alpha)
  z_target_power  <- qnorm(target_power)
  mu_orig <- z_alpha0_2 + z_p0
  mu_rep  <- z_1_minus_alpha + z_target_power
  (mu_rep / mu_orig)^2
}
```

</details>

```{r echo=FALSE, message=FALSE, warning=FALSE}

library(gt)
library(htmltools)

# Convenience wrapper: adds power for the 2.5x heuristic
compare_exact_vs_normal <- function(n_original_per_cell,
                                    target_power = 0.80,
                                    alpha = 0.05,
                                    heuristic_mult = c(".8" = 2.5, ".9" = 3.5, ".95" = 4.5),
                                    p0 = 1/3,
                                    alpha0 = 0.05) {
  if (length(heuristic_mult) > 1) {
    heuristic_mult = c("0.8" = 2.5, "0.9" = 3.5, "0.95" = 4.5)[as.character(target_power)]
  }
  approx_mult <- small_telescopes_multiplier_normal(target_power, alpha, p0, alpha0)
  map_dfr(n_original_per_cell, function(n0) {
    ex <- small_telescopes_multiplier_exact(n0, target_power, alpha, p0, alpha0)
    d33 <- ex$d33
    n_rep_heur  <- ceiling(heuristic_mult * n0)
    power_heur  <- power_reject_d33_under_null0(n_rep_heur, d33, alpha)
    bind_cols(
      ex %>% mutate(
        d33 = round(d33, 3),
        approx_multiplier = round(approx_mult, 3)
      ),
      tibble(
        heuristic_multiplier = heuristic_mult,
        n_replication_per_cell_heuristic = n_rep_heur,
        heuristic_power = power_heur
      )
    )
  }) %>%
    mutate(across(c(exact_multiplier, approx_multiplier,
                    achieved_power_exact, heuristic_power), ~round(., 3)))
}

# target-specific gt table (compact columns)
gt_power_table <- function(n0s, target_power, heuristic_mult) {
  df <- compare_exact_vs_normal(n0s, target_power = target_power, heuristic_mult = heuristic_mult) %>%
    select(n_original_per_cell, d33, exact_multiplier, achieved_power_exact, heuristic_power) %>%
    mutate(
      d33 = round(d33, 3),
      exact_multiplier = round(exact_multiplier, 3),
      achieved_power_exact = round(achieved_power_exact, 3),
      heuristic_power = round(heuristic_power, 3)
    )

  gt(df) |>
    cols_label(
      n_original_per_cell = html("n<sub>0</sub> per cell"),
      d33 = html("d<sub>33%</sub> (orig)"),
      exact_multiplier = html("exact m"),
      achieved_power_exact = html("achieved power (exact)"),
      heuristic_power = html("achieved power (heuristic)")
    ) |>
    tab_header(
      title = md(paste0("**Target power = ", 100*target_power, "%**")),
      subtitle = md(paste0("Heuristic multiplier: **", heuristic_mult, "×**"))
    ) |>
    fmt_number(columns = everything(), decimals = 2) |>
    fmt_number(columns = n_original_per_cell, decimals = 0) |>
    tab_options(
      table.font.size = px(14),
      data_row.padding = px(4),
      table.width = pct(100)
    )
}

# Render three gt tables side-by-side (80%, 90%, 95%)
render_power_gt_row <- function(n0s = c(10, 20, 50, 100, 500, 1000),
                                heuristics = c("0.8" = 2.5, "0.9" = 3.5, "0.95" = 4.5)) {
  tbl80 <- gt_power_table(n0s, 0.80, heuristics["0.8"])
  tbl90 <- gt_power_table(n0s, 0.90, heuristics["0.9"])
  tbl95 <- gt_power_table(n0s, 0.95, heuristics["0.95"])

  # Emit raw HTML to place the three gt tables in a flex row
  html <- htmltools::tagList(
    tags$div(
      style = "display:flex; gap:16px; align-items:flex-start; flex-wrap:wrap;",
      tags$div(style="flex:1 1 300px; min-width:260px;", HTML(gt::as_raw_html(tbl80))),
      tags$div(style="flex:1 1 300px; min-width:260px;", HTML(gt::as_raw_html(tbl90))),
      tags$div(style="flex:1 1 300px; min-width:260px;", HTML(gt::as_raw_html(tbl95)))
    )
  )
  knitr::asis_output(as.character(html))
}

# Example call (will output three side-by-side gt tables):
render_power_gt_row(n0s = c(10, 20, 50, 100, 500, 1000, 100000))

```

## Practical takeaway

- While the original Simonsohn (2015) heuristic of **2.5×** the original per‑cell \(n\) for **80%** power slightly underestimates the required sample size when n > 20, it achieves at least 78% power even in large samples.
- If you need **90%** or **95%** power, **3.5×** and **4.5×** can serve as comparable heuristics. They also slightly underestimate the sample size, but still achieve 89% and 94% power respectively in large samples.
- If you want to power your replication study more precisely, you can use the `small_telescopes_multiplier_exact()` function in the code above.

## Reference

Simonsohn, U. (2015). Small telescopes: Detectability and the evaluation of replication results. *Psychological Science*, 26(5), 559-569.

---
[^1] In prose we say “33% power,” but numerically we follow the Simonsohn (2015) supplement and compute with \(p_0=1/3\). Using \(p_0=.33\) instead shifts the normal multipliers slightly (e.g., 2.68 vs 2.64). 
[^2] All \(n\) here are **per‑cell** for two‑sample *t* with equal variances. Simonsohn (2015) shows that &Chi;<sup>2</sup> tests behave similarly, and various other tests directly depend on the *t*-distribution (e.g. correlations, regression coefficients). Nevertheless, the stability of the rules-of-thumb across designs could be investigated further.
